---
title: "R: Functions"
format:
  html:
    toc: true
    self-contained: true
editor_options: 
  chunk_output_type: console
---

## Functions in R

Functions in R are used to **encapsulate reusable code**. R provides a wide range of **built-in functions**, and users can also define their own functions.

---

### Built-in Functions in R

R has many **predefined functions** for mathematical operations, statistics, and data manipulation.

#### 🔹 Common Built-in Functions

```{r}
sqrt(25)
abs(-10)
round(3.14159, 2)
```


```{r}
# Character functions
toupper("hello")  # Convert to uppercase → "HELLO"
tolower("WORLD")  # Convert to lowercase → "world"
nchar("Hello")    # Count characters → 5
```

```{r}
# Logical functions
any(c(TRUE, FALSE, FALSE))   # TRUE if at least one TRUE
all(c(TRUE, TRUE, FALSE))    # FALSE if any FALSE
```



## Defining a function

Functions in R are defined using the keyword `function()`. All the statements within a function are enclosed with `{}` braces. Look at the function defined below. It takes an integer as an argument, and prints whether the integer is odd or even.


```{r}
odd_even <- function(intgr) {
  if (intgr %% 2 == 0) {
    print("even")
  } else {
      print("odd")
  }
}

odd_even(3)
```

### Function arguments

In both R and Python, functions 
support multiple types of arguments, including **positional arguments**, **default arguments**, **variable-length arguments**, and **keyword arguments**. The behavior of function arguments in R is nearly identical to Python.


### Practice exercise

Write a function that returns all prime numbers between $a$ and $b$, where $a$ and $b$ are parameters of the function.


```{r}
prime <- function(a, b) {
  prime_numbers <- c()
  for (number in a:b) {
    prime = 1
    
    for (factor in 2:(number - 1)) {
      if (number %% factor == 0) {
        prime = 0
      }
    }
    
    if (prime == 1) prime_numbers <- c(prime_numbers, number)
  }
  return(prime_numbers)
}
prime(40, 60)
```